% -*-latex-*-
% Document name: significance.tex
% Creator: Rob MacLeod [macleod@cvrti.utah.edu]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Explain the importance of the problem or critical barrier to progress in
% the field that the proposed project addresses.  Explain how the proposed
% project will improve scientific knowledge, technical capability, and/or
% clinical practice in one or more broad fields.  Describe how the
% concepts, methods, technologies, treatments, services, or preventative
% interventions that drive this field will be changed if the proposed aims
% are achieved.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Significance}
\label{sec:signif}

The most common reason for a visit to the emergency department is chest pain caused by myocardial ischemia\cite{BLZ:Saf2018,BLZ:Bhu2010}, which occurs when the perfusion to a specific region of the heart is inadequate.\cite{BMB:Hea94,BMB:Fal2007} Such inadequate perfusion, in combination with cardiac stress, leads to acute changes in the tissue and eventually lead to cell death. Well before this cell death, an insufficient supply of oxygen and other nutrients in combination with poor metabolite removal creates a toxic extracellular milieu that prevents cardiomyocytes from functioning normally.\cite{BMB:Kat2011,BMB:Foz86} These acute transient ischemic events cause the stereotypical crushing chest pain that ceases after rest, \ie{} after removal of the cardiac stress.\cite{BMB:Kat2011,BMB:Sur2011b}

A surprising range of pathologies can cause acute transient ischemia, including coronary artery disease, coronary microvascular dysfunction, Takotsubo cardiomyopathy, and coronary artery dissection.\cite{BLZ:Saf2018,BLZ:Jes2013,BLZ:Noe2017,BLZ:Jes2012} Each pathology carries a significant risk of short- and long-term mortality that can be reduced substantially by early detection.\cite{BLZ:Noe2017} \textbf{Therefore, detecting myocardial ischemia early is paramount to prevent long-term negative consequences.}\cite{BMB:Kon99,BLZ:Saf2018,BLZ:Knu2018a}

Detecting acute ischemia requires cardiac stress, which is often defined as inadequate perfusion for the cardiac metabolic demand of a region of the heart.\cite{BLZ:Pue2004} In most clinical circumstances, the safest method to induce and detect myocardial ischemia is to increase metabolic demand by having the patient exercise or by administering a pharmacological infusion to increase heart rate and contractility.\cite{RSM:Ste2002,BLZ:Saf2018,BLZ:Knu2018a}

Once the heart has been stressed, transient myocardial ischemia can be detected by several different markers. Initial elevations of extracellular potassium concentrations and anaerobic metabolism alter electrical action potentials and are followed by changes to mechanical contraction, leading to regional hypokenesis and myocardial wall thinning.\cite{BLZ:Mai1994,RSM:Rei86b} Electrical responses include a depolarization of the resting membrane potential of the cell, a decrease of the upstroke velocity of the cardiac action potential, shortening of the plateau-phase and overall duration of the action potential, and reduction in its amplitude. \cite{BLZ:Yan1996,BMB:Kat2011} The resulting spatial gradients of extracellular potentials between healthy and ischemic tissues cause intracellular current to flow between healthy and ischemic tissue, known as ``injury currents'', which can be detected from the body surface.\cite{KKA:Sha97a,KKA:Sha97b,RSM:Bra76,RSM:Bra88} These currents are largest during the ST segment of an electrocardiogram (ECG) and cause a positive or negative deflection, depending on the amount and location of the ischemic tissue and the location of the ECG electrodes. \cite{BLZ:Jan1980,BLZ:Jan1981} These electrical changes are transient and manifest only when the tissue is ischemic. This dynamic complex process of changes rapidly through an ischemic episode and creates three-dimensional zones of affected tissue with varying shapes and sizes.  Detection and localization of these ischemic zones make natural targets for diagnosis and monitoring as they have electrical consequences that can, in principle, be measured noninvasively. 

Several methods are used clinically to identify these ischemic zones. Echocardiography uses ultrasound imaging acquired during pharmacological stress to identify regions of mechanical dysfunction and wall thinning.\cite{BLZ:Man1988,BLZ:Gre1997,KKA:Gel97} The ultrasound imaging provides coarse snapshots of the ischemic region, but it is limited to discrete intervals in time and by its spatial resolution.\cite{BLZ:Saf2018} The sensitivity and specificity of stress echocardiography are 85\% and 80\%, respectively.\cite{BLZ:Knu2018a} Myocardial perfusion imaging (nuclear imaging) can also be used to detect ischemia by injecting a radioisotope during peak cardiac stress (pharmacological or exercise). Areas of decreased perfusion, which are likely ischemic, acquire a lower concentration of radioisotopes than regions with healthy perfusion. Myocardial perfusion imaging has a sensitivity and specificity of 87\% and 70\%, respectively, requires several professionals, entails a high radiation dose, and is expensive.\cite{BLZ:Knu2018a} Biochemical markers such as troponin have potential value for measuring ischemia, but they lack adequate validation and provide almost no temporal or spatial information about the ischemia induced.\cite{BLZ:Saf2018} The 12-lead electrocardiogram is perhaps the most commonly used marker during exercise heart stress and follows well the temporal changes of ischemia over time, but it provides poor spatial resolution. The sensitivity and specificity of the electrocardiographic stress test are in the range of 55--65\%.\cite{BLZ:Knu2018a,RSM:Ste2002} \textbf{We propose novel approaches that can provide both high temporal and spatial resolution and represent a functional imaging modality based on high-resolution ECG.}

The first step to improving detection of acute ischemia is to understand the electrical ischemic sources. Experimental models used previously have documented the cellular, tissue, and whole organ manifestations of myocardial ischemia.\cite{BMB:Hol77a,BMB:Hol77b,BLZ:Kle1978, RSM:Jan80,KKA:Cin80,RSM:Cha89,RSM:Mac95e,RSM:Mac97} Holland and Brooks \etal~performed the first detailed and comprehensive animal studies in the late 1970s and documented the electrical changes that occurred during myocardial ischemia.\cite{BMB:Hol77a,BMB:Hol77b}, driving much of the conventional clinical thinking, including the idea that ischemia first develops near the endocardial border and progresses radially toward the epicardium.\cite{BMB:Hol77a,BMB:Hol77b,BMB:Kat2011} Recently, our group has improved this experimental model dramatically by increasing the temporal sampling and recording electrode density in the myocardium and on the heart surface.\cite{RSM:Sho2007,RSM:Ara2009,RSM:Ara2011, BMB:Ara2015,BMB:Ara2016} Our studies have shown sharply different patterns of ischemia that initiate throughout the myocardial wall, which has motivated a complete reevaluation of the electrical sources created during ischemia.

Limits to the translational potential of these findings have included the lack of simultaneous body-surface ECGs and the absence of clinically relevant ischemia induction. In addition, no studies to our knowledge have directly compared the ischemic electrical profile using different heart stressing mechanisms. The most common comparison reported is the use of a 12-lead ECG and echocardiography to create a binary selection of either the presence or the absence of ischemia.\cite{BLZ:Sha1998a,BLZ:Dho2000,BLZ:Mai1994} Some researchers have examined the ischemic zones detected with nuclear medicine techniques, but they did not measure the electrical presence of ischemia within the heart.\cite{BLZ:San1998} \textbf{We propose to address these limitations in aims 1 and 2, the result of which will be the most detailed and translational model of acute ischemia to date. We will measure realistic ischemic sources simultaneously within the heart, on the heart surface, and on the body surface, and will also determine experimentally the role of different cardiac stress mechanisms in the development of ischemia.}

A better understanding of ischemic sources from these experiments will also help drive a better method of detection. The transient, complex, three-dimensional nature of ischemia makes it especially difficult to identify noninvasively. The 12-lead ECG provides a snapshot of the ischemic zone, if it happens to exist in a location that is monitored by one of the leads. An imaging-like approach is an attractive and potentially powerful option to capture the dynamic and complex nature of ischemic sources.

This imaging is known as ECG Imaging (ECGI)\cite{RSM:Bro97b,RSM:Pul2010,RSM:Rud2015}; analogous to all imaging modalities, it uses individual electrodes to capture single views of the electrical potential in the heart and reconstructs them to capture cardiac electrical activity from multiple directions.  This process is much like a series of X-rays taken from different angles and then reconstructed to create a three-dimensional CT scan. ECGI is an approach with several decades of research that has recently been used in a various clinical applications, including premature ventricular contraction localization, atrial fibrillation detection, and other electrical abnormalities.\cite{RSM:Sha2015,BLZ:Pot2014,BLZ:Dub2015,BLZ:Wan2016a,RSM:Sha2015,BLZ:Wan2018,RSM:Ost97,RSM:Mac98,RSM:Bro99,RSM:Clu2018,RSM:Clu2015} This approach can provide a noninvasive, continuous, and patient-specific view of the electrical activity in the heart and thus a functional status that could dramatically improve diagnostic accuracy and monitoring capability. As ECGI specialists, we have designed open-source toolkits that have been applied to a range of pathologies, including acute ischemia.\cite{BMB:Mac95,BMB:Bur2011,BMB:Bur2018a,BMB:Bur2018b,BLZ:Tat2018} To provide enough information for an imaging approach, more electrodes in more locations are required than the standard 12-lead ECG. This approach has been known for several decades as ``body-surface potential mapping'' or BSPM, and it provides a complete, but remote, sampling of the myocardial electrical potentials. \cite{BLZ:Mil1980,BLZ:Fox1979,BLZ:Hor2001,RSM:Koz95}

Despite the potential utility and the clinical need, ECGI has rarely been applied or validated in the setting of acute transient myocardial ischemia. Other groups have used ECGI to study ischemia and conducted validation studies using clinical imaging. Nielsen \etal~acquired SPECT cardiac stress testing images and body-surface potentials from four patients undergoing a standard exercise stress test.\cite{BMB:Nie2013} They used used ECGI to reconstruct the ischemic zones at the limited resolution of the 17-segment model of the ventricles and further classified each zone as endocardial, epicardial, or transmural, findings they confirmed with cardiac SPECT. The limitations of this study included the absence of electrical confirmation and the very low time resolution inherent in SPECT. Similarly, the relatively low resolution of their approach (centimeters rather than millimeters) hindered the analysis of the complex ischemic zones that arise and deform rapidly during cardiac stress.

 Few previous studies have attempted to implement ECGI as a viable method to detect myocardial ischemia. The first was from the mentor of this training grant \cite{RSM:Mac95} and while it showed promising results, these were limited to epicardial potentials and lacked consistency.  More recently, members of our group have created and applied a fundamentally novel mathematical formulation based on a new source model to localize the ischemic source within the myocardium. This advance also required novel use of numerical solvers that created very promising framework to detect acute myocardial ischemia at the level of the transmembrane potential (TMP).\cite{RSM:Wan2013,RSM:Wan2011a} The limitation of these studies is that they used simulated datasets and manually imposed `ischemic zones' is a myocardial geometry, from which they generated body-surface potentials at a single time instant to form the input data for the ECGI. \textbf{These methods showed promise but were limited by the lack of ground truth datasets from experiments of suitable quality. With our expertise and recent improvements to the experimental model, we are perfectly poised to implement novel detection techniques to identify ischemic sources in the heart. Accomplishing aim 3 would be significant because it would result in a new comprehensive and accurate method to reconstruct the spatiotemporal image of acute myocardial ischemia.}

In summary, this project is significant for several reasons. First, the novel experimental preparation will be a first of its kind, with high-resolution recordings and transient ischemic control that maximize the potential to translate any findings into clinical practice. Second, we will compare and contrast different, clinically relevant types of ischemic stress in a highly instrumented and physiologically relevant animal preparation to better detect differences that could never before be seen. Finally, we will apply novel ECGI techniques to improve the detection and localization of ischemia, which will better inform physicians of the progression and status of acute ischemia in a patient-specific manner and thus improve the unacceptably poor performance of current clinical methods.

